---
output: github_document
author: Barry Grant < http://thegrantlab.org >
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
# Demo1: R Language Basics

### Background

R is powerful data programming language that you can use to explore and understand data in an open-ended, highly interactive, iterative way. Learning R will give you the freedom to experiment and problem solve during data analysis — exactly what we need as bioinformaticians.

Before delving into working with real data in R, we need to learn the basics of the R language. In this section, we’ll learn how to do simple calculations in R, assign values to variables, and call functions. Then, we’ll look at R’s vectors, vector data types, and vectorization. Vectors and vectorization underpin how we approach many problems in R.

### Simple Calculations

Open RStudio, or a terminal R prompt, and try typing the basic arithmetic below. If you are using RStudio we will be typing directly into the *Console* that contains the live R process.

```{r}
5+3
5-3
5*3
5/3
```

Lets make an assignment and then inspect the object you just created.  

```{r}  
x <- 3 * 4  
x
```

All R statements where you create objects -- "assignments" -- have this form:

```{r, eval=FALSE}
objectName <- value  
```

and in my head I hear, e.g., "x gets 12".

You will make lots of assignments and the operator `<-` is a pain to type. Don't be lazy and use `=`, although it would work, because it will just sow confusion later. Instead, utilize RStudio's keyboard shortcut: Alt + - (the minus sign).

Notice that RStudio automagically surrounds `<-` with spaces, which demonstrates a useful code formatting practice. Code is miserable to read on a good day. Give your eyes a break and use spaces.

RStudio offers many handy [keyboard shortcuts](https://support.rstudio.com/hc/en-us/articles/200711853-Keyboard-Shortcuts). Also, Alt+Shift+K brings up a keyboard shortcut reference card.

Object names cannot start with a digit and cannot contain certain other characters such as a comma or a space. You will be wise to adopt a [convention for demarcating words](http://en.wikipedia.org/wiki/Snake_case) in names.

```{r, eval=FALSE}
i_use_snake_case  
other.people.use.periods  
evenOthersUseCamelCase  
```

Make another assignment

```{r}
this_is_a_really_long_name <- 2.5  
```

To inspect this, try out RStudio's completion facility: type the first few characters, press TAB, add characters until you disambiguate, then press return.

Make another assignment
```{r}
r_rocks <- 2 ^ 3  
```

Let's try to inspect:

```{r, eval=FALSE}
rrocks
```

```
## Error in eval(expr, envir, enclos): object 'rrocks' not found
```

What happened? What about this one?

```{r, eval=FALSE}
R_rocks
```

> **Conclusion:** When we programm we enter into an implicit contract with the computer. The computer will do tedious computation for you. In return, you will be completely precise in your instructions. Typos matter. Case matters. **Therefore get better at typing!**


### Calling Functions
R has a mind-blowing collection of built-in functions that are accessed like so:


```{r, eval=FALSE}
functionName(arg1 = val1, arg2 = val2, and so on)
```

Let's try using the `seq()` function, which makes regular sequences of numbers and, while we're at it, demo more helpful features of RStudio.

Type `se` and hit TAB. A pop up shows you possible completions. Specify `seq()` by typing more to disambiguate or using the up/down arrows to select. 

Notice the floating tool-tip-type help that pops up, reminding you of a function's arguments. If you want even more help, press F1 as directed to get the full documentation in the help tab of the lower right pane. Now open the parentheses and notice the automatic addition of the closing parenthesis and the placement of cursor in the middle.  Type the arguments `1, 10` and hit return. RStudio also exits the parenthetical expression for you. IDEs are helpful when starting out!

```{r}
seq(1,10)  
```

You can always access the help/documentation of a particular function by using the `help(myFunction)` function, e.g. `help(seq)`.

The previous `seq()` example above also demonstrates something about how R resolves function arguments. You can always specify in `name = value` form. But if you do not, R attempts to resolve by position. So above, it is assumed that we want a sequence `from = 1` that goes `to = 1`. Since we didn't specify step size, the default value of `by` in the function definition is used, which ends up being 1 in this case. For functions I call often, I might use this resolve by position for the first
argument or maybe the first two. After that, I always use `name = value`.


What does the following command do?
```{r}
seq(1,10, by=2)  
```

Could you of figured this out from the documentation accessed via the command `?seq` or `help(seq)`? One very useful part of a functions documentation is the **Examples** section (typically found at the very end of the documentation entry).

You can copy and paste these into your session or execute them all in one go with the `examples()` function, e.g.

```{r, eval=FALSE}
example(seq)
```

Not all functions have (or require) arguments, e.g.:

```{r}
date()
```

### Getting help in R
As would be expected from a sophisticated scientific and statistical computing language, R has oodles of functions—far more than any reasonable human can expect to learn and remember. Consequently, you’ll need to master two of the most practical R skills:

  * Knowing how to look up a function’s documentation to recall its arguments and how it works.

  * Being able to discover new useful functions.
  
Each of R’s functions (and other objects such as constants like pi, classes, and packages) has integrated documentation accessible within R. R’s documentation includes descriptions and details about the function, all arguments of a function, and useful usage examples. You access R’s built-in documentation with the `help()` function or its syntactic shortcut, `?`:

```{r, eval=FALSE}
help(log)
?log
```

In RStudio, this opens a special help window containing log()’s documentation. In terminal R, this documentation is handled by your default pager (probably the program less). Operators such as + and ^ need to be quoted (e.g., help('+')). 

R’s `help()` function is useful when you already know the name of the function you need documentation for. Unfortunately, we often only have a fuzzier idea of what we need help with (e.g., what was the function in R that calculates cross tabulate vectors?). For tasks like this, we can search R’s help system with the function `help.search()`, or its shortcut `??`:

```{r, eval=FALSE}
help.search("cross tabulate")
??"cross tabulate"
```

In this case, `help.search()` would help you find the function `table()`, which is useful in creating counts and cross tabulations from vectors.

Also, R has the neat feature that all examples in an R help file can be executed with the function `example()`. For example:

```{r}
example(log)
```

R also has functions for listing all functions in a package (e.g., `library(help="base")`) and finding functions by name (e.g., `apropos(norm)`), which are often useful in remembering a function’s name. 

Later today we will cover a special form of package documentation called **vignettes**. These are basically short task oriented introductions and tutorials for a particular set of related functions.


### Vectors, Vectorization, and Indexing
Arguably the most important feature of the R language is its vectors. A vector is a container of contiguous data. Unlike most languages, R does not have a type for a single value (known as a scalar) such as 3.1 or “AGCTACGACT.” Rather, these values are stored in a vector of length 1. We can verify that values like 3.1 are vectors of length 1 by calling the function length() (which returns the length of a vector) on them:

```{r}
length(3.1) 
```

To create longer vectors, we combine values with the function c():

```{r}
x <- c(56, 95.3, 0.4)
x
y <- c(3.2, 1.1, 0.2)
y
```

#### Vectorization
R’s vectors are the basis of one of R’s most important features: **vectorization**. Vectorization allows us to loop over vectors elementwise, without the need to write an explicit loop. For example, R’s arithmetic operators are all vectorized:

```{r}
x+y
x-y
x/y
```

In addition to operators like `+` and `*`, many of R’s math functions (e.g., `sqrt()`, `round()`, `log()`, etc.) are all vectorized:

```{r}
sqrt(x)
round(sqrt(x), 3)
log(x)/2 + 1 # note how we can combined vectorized operations
```

This vectorized approach is not only more clear and readable, it’s also often computationally faster.

> **Concludion:** Unlike other languages, R allows us to completely forgo explicitly looping over vectors with a for loop. Later on, we’ll see other methods used for more explicit looping.

#### Vector Indexing (or getting to your data subset)
We can access specific elements of a vector through indexing. An index is an integer that specifies which element in the vector to retrieve. We can use indexing to get or set values to certain elements from a vector:

```{r}
x <- c(56, 95.3, 0.4)
x[2]
```

R’s vectors are 1-indexed, meaning that the index 1 corresponds to the first ele‐ ment in a list (in contrast to 0-indexed languages like Python). Here, the value 95.3 is the 2nd item, and is accessed with x[2].

```{r}
x[1]
x[4]
```
Trying to access an element that doesn’t exist in the vector leads R to return NA, the “not available” missing value.

```{r}
x[3] <- 0.5
x
```
We can change specific vector elements by combining indexing and assignment.


Our next session will cover more on **vectors**, and the other major R data structures (**matrices**, **data.frames** and **lists**).


### A note about workspace and working directory

One day you will need to quit R, go do something else and return to your analysis later.

One day you will have multiple analyses going that use R and you want to keep them separate.

One day you will need to bring data from the outside world into R and send numerical results and figures from R back out into the world.

To handle these real life situations, you need to make two decisions:

  * What about your analysis is "real", i.e. you will save it as your lasting record of what happened?

  * Where does your analysis "live"?

#### Workspace, .RData

As a beginning R user, it's OK to consider your workspace "real". _Very soon_, I urge you to evolve to the next level, where you consider your saved R scripts as "real". (In either case, of course the input data is very much real and requires preservation!) With the input data and the R code you used, you can reproduce
_everything_. You can make your analysis fancier. You can get to the bottom of puzzling results and discover and fix bugs in your code. You can reuse the code to conduct similar analyses in new projects. You can remake a figure with different aspect ratio or save is as TIFF instead of PDF. You are ready to take questions. You are ready for the future.

If you regard your workspace as "real" (saving and reloading all the time), if you need to redo analysis ... you're going to either redo a lot of typing (making mistakes all the way) or will have to mine your R history for the commands you used. Rather than [becoming an expert on managing the R history](https://support.rstudio.com/hc/en-us/articles/200526217-Command-History), a better use of your time and psychic energy is to keep your "good" R code in a script for future reuse.

Because it can be useful sometimes, note the commands you've recently run appear in the History pane.

But you don't have to choose right now and the two strategies are not incompatible. Let's demo the save / reload the workspace approach.

Upon quitting R, you have to decide if you want to save your workspace, for potential restoration the next time you launch R. Depending on your set up, R or your IDE, e.g. RStudio, will probably prompt you to make this decision.

Quit R/Rstudio, either from the menu, using a keyboard shortcut, or by typing `q()` in the Console. You'll get a prompt like this:

> Save workspace image to ~/.Rdata?

_Note where the workspace image is to be saved_ and then click `Save`.

Using your favorite method, visit the directory where image was saved and verify there is a file named `.RData`. You will also see a file `.Rhistory`, holding the commands submitted in your recent session.

Restart RStudio. In the Console you will see a line like this:

```
[Workspace loaded from ~/.RData]
```

indicating that your workspace has been restored. Look in the Workspace pane and you'll see the same objects as before. In the History tab of the same pane, you should also see your command history.You're back in business. This way of starting and stopping analytical work will not serve you well for long but it's a start.

#### Working directory

Any process running on your computer has a notion of its "working directory". In R, this is where R will look, by default, for files you ask it to load. It also where, by default, any files you write to disk will go. Chances are your current working directory is the directory we inspected above, i.e. the one where RStudio wanted to save the workspace.

You can explicitly check your working directory with:


```{r, eval=FALSE}
getwd()
```

It is also displayed at the top of the RStudio console.

As a beginning R user, it's OK let your home directory or any other weird directory on your computer be R's working directory. _Very soon_, I urge you to evolve to the next level, where you organize your analytical projects into directories and, when working on project A, set R's working directory to the associated directory.

__Although I do not recommend it__, in case you're curious, you can set R's working directory at the command line like so:


```{r, eval=FALSE}
setwd("~/myCoolProject")
```

__Although I do not recommend it__, you can also use RStudio's Files pane to navigate to a directory and then set it as working directory from the menu: Session --> Set Working Directory --> To Files Pane Location. (You'll see even more options there). Or within the Files pane, choose __More__ and __Set As
Working Directory__.

But there's a better way. A way that also puts you on the path to managing your R work like an expert.

### RStudio projects

Keeping all the files associated with a project organized together -- input data, R scripts, analytical results, figures -- is such a wise and common practice that RStudio has built-in support for this via its _projects_.

[Using Projects](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects)

Let's make one to use for the rest of this class. Do this: File > New Project .... The directory name you choose here will be the project name. Call it whatever you want (or follow me for convenience *Desktop/bootcamp/R/session1*).

### Miscellaneous notes

It is traditional to save R scripts with a `.R` or `.r` suffix. Follow this convention unless you have some extraordinary reason not to. 

Comments start with one or more `#` symbols. Use them. RStudio helps you (de)comment selected lines with Ctrl+Shift+C (windows and linux) or Command+Shift+C (mac).

Clean out the workspace, ie pretend like you've just revisited this project after a long absence.  The broom icon or `rm(list = ls())`. Good idea to do this, restart R (available from the Session menu), re-run your analysis to truly check that the code you're saving is complete and correct (or at least rule out obvious problems!).

This work-flow will serve you well in the future:

  * Create an RStudio project for an analytical project
  * Keep inputs there (we'll soon talk about importing)
  * Keep scripts there; edit them, run them in bits or as a whole from there
  * Keep outputs there (like the PDF written above)

Avoid using the mouse for pieces of your analytical work-flow, such as loading a dataset or saving a figure. Terribly important for reproducibility and for making it possible to retrospectively determine how a numerical table or PDF was actually produced (searching on local disk on filename, among `.R` files, will lead to the relevant script).

Many long-time users never save the workspace, never save `.RData` files (I'm one of them), never save or consult the history. Once/if you get to that point, there are options available in RStudio to disable the loading of .RData and permanently suppress the prompt on exit to save the workspace (go to Tools->Options->General).

For the record, when loading data into R and/or writing outputs to file, you can always specify the absolute path and thereby insulate yourself from the current working directory. This is rarely necessary when using RStudio Projects properly

#### Reproducibility and sessionInfo()
Versions of R and any R packages installed change over time. This can lead to reproducibility headaches, as the results of your analyses may change with the changing version of R and R packages. Solving these issues is an area of ongoing development (see, for example, the packrat package). At the very least, you should always record the versions of R and any packages you use for an analysis. R actually makes this incredibly easy to do—just call the `sessionInfo()` function:

```{r}
sessionInfo()
```
